LabMate: a prospectus for types for MATLAB

### 2024-04-10 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] \mathit FTW

Under the hood, LabMate translates MATLAB expressions to typed
expressions in its own core type theory.
%
Since matrices feature heavily in MATLAB code, the type of matrices
plays a central role in this type theory.
%
We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we parameterise the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $rs$ and the $j$th entry
in the list $cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Matrix types formation can thus be summarised as the inference rule
given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
      {
        \istype{R}
        &\quad
        \istype{C}
        &\quad
        x : R, y : C \vdash \istype{E(x,y)}
        &\quad
        rs : \List{R}
        &\quad
        cs : \List{C}
      }
    \]
  \end{center}
  \caption{Type formation rule for matrix types}
  \label{fig:matrix_intro}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
      {
        x : R, y : C \vdash \isadd{E(x, y)}
        &\quad
        A : \Matrix{R}{C}{E}{rs}{cs}
        &\quad
        B : \Matrix{R}{C}{E}{rs}{cs}
      }
    \]
  \end{center}
  \caption{Addition rule for matrices}
  \label{fig:matrix_add}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{rs}{ms}
        &\quad
        B : \Matrix{R}{C}{E'}{ms}{cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}

A notable special case is when both $R$ and $C$ are the unit type
$R = C = \One$, that is, the type with exactly one element.
%
In this case, there is only one type of entries $E$ (because both $x$
and $y$ must be the unique element of the unit type), and the only
information available in the lists $rs : \List{\One}$ and
$cs : \List{\One}$ is their length.
%
Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
$m \times n$ matrices with entries of type $E$, where $n$ and $m$ are
the unique lists over the unit type of length $n$ and $m$
respectively.

Another noteworthy special case which we want to support takes $R$ and
$R$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
type of physical quantities of dimension $d_1 \cdot d_2^{-1}$.
%
This way, we can reduce dimensional consistency checking \`a la
Kennedy~\cite{kennedyUOM} also for programs involving matrices,
following the work of Hart~\cite{hart}.
%
We will see in Section~\ref{sec:example-revisited} how this can be helpful.

Terms of matrix type in the core type theory are built up from
$1 \times 1$ matrices and pasting matrices horizontally and vertically.
%
Singleton matrices have the following typing rule
\[
  \infer{[e] : \Matrix{R}{C}{E}{[r]}{[c]}}
    {r : R
      &\quad
      c : C
      &\quad
      e : E(r, c)
    }
\]
%
while horizontal and vertical pasting horizontally have the following
typing rules:
\[
  \infer{\hjux{A}{B} : \Matrix{R}{C}{E}{rs}{(cs_A \append cs_B)}}
  {
    A : \Matrix{R}{C}{E}{rs}{cs_A}
    &\quad
    B : \Matrix{R}{C}{E}{rs}{cs_B}
  }
\]
\[
  \infer{\vjux{A}{B} : \Matrix{R}{C}{E}{(rs_A \append rs_B)}{cs}}
  {
    A : \Matrix{R}{C}{E}{rs_A}{cs}
    &\quad
    B : \Matrix{R}{C}{E}{rs_B}{cs}
  }
\]
When comparing matrices for equality, we reassociate horizontal and
vertical pastings, so that $\hjux{\vjux{A}{C}}{\vjux{B}{D}}$ and
$\vjux{\hjux{A}{B}}{\hjux{C}{D}}$ indeed are considered equal.

>>>>

\newcommand{\rs}{\mathit{rs}}
\newcommand{\ms}{\mathit{ms}}
\newcommand{\cs}{\mathit{cs}}

Under the hood, LabMate translates MATLAB expressions to typed
expressions in its own core type theory.
%
Since matrices feature heavily in MATLAB code, the type of matrices
plays a central role in this type theory.
%
We make this type rather precise, by allowing the type of each entry
in the matrix to vary according to the row and column of the entry:
%
we parameterise the type of matrices by a type $R$ associated with
each row, a type $C$ associated with each column, and a type $E(x,y)$
which might vary depending on the values of parameters $x : R$ and
$y : C$, as well as two lists $\rs : \List{R}$ and $\List{C}$ --- the
idea being that the $i$th entry in the list $\rs$ and the $j$th entry
in the list $\cs$ determines the type of the $(i, j)$ coordinate in the
matrix.
%
Matrix types formation can thus be summarised as the inference rule
given in Figure~\ref{fig:matrix_intro}.
%\[
%  \infer{\istype{\Matrix{R}{C}{E}{rs}{\cs}}}
%  {
%    \istype{R}
%    &\quad
%    \istype{C}
%    &\quad
%    x : R, y : C \vdash \istype{E(x,y)}
%    &\quad
%    rs : \List{R}
%    &\quad
%    cs : \List{C}
%  }
%\]
%
\begin{figure*}[th]
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
      {
        \istype{R}
        &\quad
        \istype{C}
        &\quad
        x : R, y : C \vdash \istype{E(x,y)}
        &\quad
        \rs : \List{R}
        &\quad
        \cs : \List{C}
      }
    \]
  \end{center}
  \caption{Type formation rule for matrix types}
  \label{fig:matrix_intro}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
      {
        x : R, y : C \vdash \isadd{E(x, y)}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\cs}
        &\quad
        B : \Matrix{R}{C}{E}{\rs}{\cs}
      }
    \]
  \end{center}
  \caption{Addition rule for matrices}
  \label{fig:matrix_add}
\end{subfigure}
\hfill
\begin{subfigure}{\textwidth}
  \begin{center}
    \[
      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
      {
        \ismult{E}{E'}{E''}
        &\quad
        A : \Matrix{R}{C}{E}{\rs}{\ms}
        &\quad
        B : \Matrix{R}{C}{E'}{\ms}{\cs}
      }
    \]
  \end{center}
  \caption{Multiplication rule for matrices}
  \label{fig:matrix_mul}
\end{subfigure}
% FIXME : \caption is broken
\captionwide{Typing rules for matrices}
\label{fig:matrix_rules}
\end{figure*}

A notable special case is when both $R$ and $C$ are the unit type
$R = C = \One$, that is, the type with exactly one element.
%
In this case, there is only one type of entries $E$ (because both $x$
and $y$ must be the unique element of the unit type), and the only
information available in the lists $\rs : \List{\One}$ and
$\cs : \List{\One}$ is their length.
%
Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
$m \times n$ matrices with entries of type $E$, where $n$ and $m$ are
the unique lists over the unit type of length $n$ and $m$
respectively.

Another noteworthy special case which we want to support takes $R$ and
$C$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
type of physical quantities of dimension $d_1 \cdot d_2^{-1}$.
%
This way, we can reduce dimensional consistency checking \`a la
Kennedy~\cite{kennedyUOM} also for programs involving matrices,
following the work of Hart~\cite{hart}.
%
We will see in Section~\ref{sec:example-revisited} how this can be helpful.

Terms of matrix type in the core type theory are built up from
$1 \times 1$ matrices and pasting matrices horizontally and vertically.
%
Singleton matrices have the following typing rule
\[
  \infer{[e] : \Matrix{R}{C}{E}{[r]}{[c]}}
    {r : R
      &\quad
      c : C
      &\quad
      e : E(r, c)
    }
\]
%
while horizontal and vertical pasting horizontally have the following
typing rules:
\[
  \infer{\hjux{A}{B} : \Matrix{R}{C}{E}{\rs}{(cs_A \append cs_B)}}
  {
    A : \Matrix{R}{C}{E}{\rs}{cs_A}
    &\quad
    B : \Matrix{R}{C}{E}{\rs}{cs_B}
  }
\]
\[
  \infer{\vjux{A}{B} : \Matrix{R}{C}{E}{(\rs_A \append \rs_B)}{cs}}
  {
    A : \Matrix{R}{C}{E}{\rs_A}{cs}
    &\quad
    B : \Matrix{R}{C}{E}{\rs_B}{cs}
  }
\]
When comparing matrices for equality, we reassociate horizontal and
vertical pastings, so that $\hjux{\vjux{A}{C}}{\vjux{B}{D}}$ and
$\vjux{\hjux{A}{B}}{\hjux{C}{D}}$ indeed are considered equal.
